---
title: "Introduction to `HermesData()`"
date: "Last updated on `r Sys.Date()`"
output: rmarkdown::html_vignette

vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---

## The `HermesData`

* The `gSummarizedExperiment()` takes `assays`, `rowData`, `colData` and `metadata` as input, while `HermesData()` should directly take `SummarizedExperiment` object as input, similar like `se2gSE()`.

* The first assay must be `counts` containing non-missing, integer, non-negative values.

* Similar to the `gSummarizedExperiment()`, `HermesData()` must have mandatory `rowData` (HGNC, HGNCGeneName, etc) and `colData` (Filename, SampleID, etc) columns. If the input does not include these `rowData` or `colData` columns, the function will fill the columns with `NA` values. 

Note: At this point, we don't yet look up gene annotation data. It will be updated later.
  
  
## Idea: take SE as argument

Let's say we need to have an assay in the SE. Otherwise we don't have any data.

First: what is if there is no `colData`? Let's try with a very simple SE:

```{r}
library(SummarizedExperiment)
se <- SummarizedExperiment(matrix(1:4, 2, 2))
se
rownames(colData(se))
```

OK, if the assay did not have row names, the SE also does not have.
But if the assay has row names:

```{r}
my_assay <- matrix(1:4, 2, 2, dimnames = list(c("a", "b"), c("c", "d")))
rownames(my_assay)
se <- SummarizedExperiment(my_assay)
rownames(colData(se))
rownames(rowData(se))
```
Then the SE also has row names.

So if we take an SE, it will already have correct row names for `colData` and `rowData.` We can see what goes wrong when using the internal constructor:

```{r}
# hermes:::.HermesData(se)
```

OK so what we want to help the user is to add the mandatory columns for `colData` and `rowData`, as well as annotate the genes (i.e. adding row data) via IGIS (we'll implement that in the next sprint).

Question: there are some columns can't be completely `NA`. These are captured in the internal package constants `.row_data_non_empty_cols` and `.col_data_non_empty_cols`. We don't try to impute these in the constructor since later this would anyway fail the validation. That means these columns need to be supplied by the input object already.

```{r}
annotate_igis <- identity # to be implemented later

HermesData <- function(object) {
  assert_that(
    is_class(object, "SummarizedExperiment"),
    not_empty(assays(object)),
    not_empty(rowData(object)),
    not_empty(colData(object))
  )
  
  object <- annotate_igis(object)
  
  missing_row <- setdiff(.row_data_additional_cols, names(rowData(object)))
  rowData(object)[, missing_row] <- NA
  
  missing_col <- setdiff(.col_data_additional_cols, names(colData(object)))
  colData(object)[, missing_col] <- NA
  
  .HermesData(object)
}
```

We don't tinker with the assays here since that needs to be already correctly set in the object (counts assay with non-negative integers etc.).
