---
title: "Design: Joining data frame with colData and gene info from `HermesData`"
author: "Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
  
## Objective

Starting point is that we have a data frame (e.g. CDISC format) containing keys (e.g. a subject ID column
`USUBJID`). We also have a `HermesData` object containing RNA-seq data, here the `colData`
needs to contain the same keys (e.g. `USUBJID`) as well. For specified genes or a gene signature
we now want to join the corresponding RNA-seq data from a specific assay to the data frame so that it
contains these as additional columns.

Note that the `HermesData` object here might have come from a larger MAE originally and now contains 
the required `colData`, but that can be done as a separate step.

## Idea

We can proceed in steps here:

1) obtain gene columns via `GeneSpec` extraction from assay as a data frame, and 
2) join that with `colData` of the experiment
3) join that to the data frame

## Prototype

### Get gene data frame

This will go as a new extraction method into the `GeneSpec` class.

```{r}
get_gene_data <- function(assay_matrix, genes) {
  gene_matrix <- genes$extract(assay_matrix)
  if (!is.matrix(gene_matrix)) {
    gene_matrix <- t(gene_matrix)
  }
  num_genes <- nrow(gene_matrix)
  gene_names <- if (num_genes == 1) {
    genes$get_label()
  } else {
    genes$get_gene_labels()
  }
  gene_names <- make.names(gene_names)
  rownames(gene_matrix) <- gene_names
  data.frame(t(gene_matrix))
}
```

Let's try it out:

```{r}
res <- get_gene_data(
  counts(hermes_data),
  gene_spec(genes(hermes_data)[1:5])
)

head(res)
```

### Get `colData` with genes

```{r}
#' @param object (`AnyHermesData`)\cr input experiment.
#' @param assay_name (`string`)\cr which assay to use.
#' @param genes (`GeneSpec`)\cr which genes or which gene signature should be extracted.
col_data_with_genes <- function(object,
                                assay_name,
                                genes) {
  assert_class(object, "AnyHermesData")
  assert_string(assay_name)
  assert_class(genes, "GeneSpec")
  
  col_data <- colData(object)
  assay_matrix <- assay(object, assay_name)
  gene_data <- get_gene_data(assay_matrix, genes)
  
  assert_true(identical(rownames(col_data), rownames(gene_data)))
  structure(
    cbind(
      col_data,
      gene_data
    ),
    gene_cols = names(gene_data)
  )
}
```

Let's try it out:

```{r}
res <- col_data_with_genes(
  hermes_data, 
  "counts", 
  gene_spec("GeneID:11185")
)

head(res)
attributes(res)$gene_cols
```

### Join that with data frame

Finally in this function we require the patient (`USUBJID`) column to be present.
Also additional keys can be specified. (TODO)

```{r}
join_cdisc <- function(cdisc_data,
                       object,
                       assay_name,
                       genes) {
  assert_data_frame(cdisc_data)
  col_data <- col_data_with_genes(
    object,
    assay_name,
    genes
  )  
  assert_names(names(cdisc_data), must.include = "USUBJID")
  assert_names(names(col_data), must.include = "USUBJID")
  
  # Check on patients.
  gene_patients <- unique(col_data$USUBJID)
  cdisc_patients <- unique(cdisc_data$USUBJID)
  patients_not_in_cdisc <- setdiff(gene_patients, cdisc_patients)
  if (length(patients_not_in_cdisc) > 0) {
    warning(paste(
      "Patients", paste(patients_not_in_cdisc, collapse = ", "),
      "removed because not contained in CDISC data set"
    ))
  }
  
  # Inner join by USUBJID.
  cols_to_take_from_col_data <- setdiff(names(col_data), "USUBJID")
  cdisc_data <- adtte[, - which(names(cdisc_data) %in% cols_to_take_from_col_data)]
  result <- merge(cdisc_data, col_data, by = "USUBJID")
}
```

### Separate check function?

I am not sure if we want a separate checking function. This could maybe stay in `tmh`.

```{r}
check_patient_id_mae <- function(mae) {
  mae_samplemap <- MultiAssayExperiment::sampleMap(mae)
  mae_coldata <- MultiAssayExperiment::colData(mae)
  sm_usubjid <- as.character(merge_samplemap$USUBJID)
  if ("USUBJID" %in% colnames(mae_coldata)) {
    mae_usubjid <- as.character(mae_coldata$USUBJID)
    assert_subset(
      x = sm_usubjid,
      choices = mae_usubjid
    )
  }
  
  # todo loop over all experiments? or only for one experiment?
  samplemap_experiment <- mae_samplemap[mae_samplemap$assay == experiment_name, ]
  patients_in_experiment <- samplemap_experiment$primary
  assert_character(patients_in_experiment, unique = TRUE)

  merge_samplemap <- samplemap_experiment[, c("primary", "colname")]
  merge_samplemap <- as.data.frame(merge_samplemap)
  colnames(merge_samplemap) <- c("USUBJID", "SampleID")

  hd <- suppressWarnings(MultiAssayExperiment::getWithColData(mae, experiment_name))
  assert_class(hd, "AnyHermesData")
  hd_usubjid <- as.character(SummarizedExperiment::colData(hd)$USUBJID)
  
  assert_subset(
    x = hd_usubjid,
    choices = sm_usubjid
  )
}
```

