---
title: "Design: Correlation between Samples and Heatmap Plot"
author: "Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This vignette explores the calculation of correlation between samples and the
resulting heatmap plot.

## Idea

- One function `cor()` should calculate the correlations between samples.
- The other function `draw_cor_heatmap()` takes those results and creates the plot.
- The advantage of this modular design is that we can better test the
correlation calculation and the user can also modify things / have their
function before putting into the drawing function.

## Annotation

Let's explore the annotation needed for the heatmap.

```{r}
object <- hermes_data
anno <- data.frame(colData(object)[c("TechnicalFailureFlag", "LowDepthFlag")])
head(anno)
anno.char <- as.data.frame(lapply(anno, function(x) ifelse(is.na(x), "NA", as.character(x))))
rownames(anno.char) <- rownames(anno)
head(anno.char)
```

So effectively this just gets the two flag variables, converts them to character with explicit missings.
We could maybe use the tern function for that:

```{r}
tern::explicit_na(as.character(c(FALSE, NA)))
```

## Correlation Calculation

Many users will know already the `cor` function, therefore the idea is to set that as a generic function
and define a method for our `hermes` objects. We not only return the correlation matrix, but as an 
attribute `flag_data` also the two flags for the samples. This will be later used for the heatmap
annotation.

```{r}
setGeneric("cor")

.HermesDataCor <- setClass(
  Class = "HermesDataCor",
  contains = "matrix",
  slots = c(flag_data = "DataFrame")
)

setMethod(
  f = "cor",
  signature = signature(x = "AnyHermesData"),
  definition = function(x, y = "counts") {
    assert_that(is.string(y))
    
    chosen_assay <- assay(x, y)
    sample_cor_matrix <- stats::cor(chosen_assay, method = "pearson")
    
    .HermesDataCor(
      sample_cor_matrix,
      flag_data = colData(x)[, c("TechnicalFailureFlag", "LowDepthFlag")]
    )
  }
)
```

Note that we cannot use the argument name `assay` instead of `y` here since the original function
does not take `...` arguments.

Let's try it out:

```{r}
result <- cor(x = object)
result[1:3, 1:3]
result@flag_data
class(result)
```

Note that we give the result an S4 class, just that we can check in the
downstream plotting function that it comes from here. Actually... then we can
just define a `plot` method to make it even more easy. We could also use S3
classes, but then it gets confusing in the package where we otherwise use S4
classes.

## Heatmap Plotting

Fortunately we already have `ComplexHeatmap` in the docker. Also fortunately there is a translation
from the previously used `pheatmap` arguments to this, see [here](https://jokergoo.github.io/ComplexHeatmap-reference/book/integrate-with-other-packages.html#translation).

```{r}
library(ComplexHeatmap)

left_annotation <- rowAnnotation(
  LowDepthFlag = factor(anno$LowDepthFlag),
  col = list(LowDepthFlag = c("FALSE" = "green", "TRUE" = "red"))
)

top_annotation <- HeatmapAnnotation(
  TechnicalFailureFlag = factor(anno$TechnicalFailureFlag),
  col = list(TechnicalFailureFlag = c("FALSE" = "green", "TRUE" = "red"))
)

Heatmap(
  matrix = result,
  col = circlize::colorRamp2(c(0, 0.5, 1), c("red", "yellow", "green")),
  name = "Correlation",
  left_annotation = left_annotation,
  top_annotation = top_annotation
)
```

Now we can define the plot method for the correlation object:

```{r}
setMethod(
  f = "plot",
  signature = "HermesDataCor",
  definition = function(x, 
                        flag_colors = c("FALSE" = "green", "TRUE" = "red"),
                        cor_colors = circlize::colorRamp2(c(0, 0.5, 1), c("red", "yellow", "green")),
                        ...) {
    df <- x@flag_data
    left_annotation <- rowAnnotation(
      LowDepthFlag = factor(df$LowDepthFlag),
      col = list(LowDepthFlag = flag_colors)
    )
    top_annotation <- HeatmapAnnotation(
      TechnicalFailureFlag = factor(df$TechnicalFailureFlag),
      col = list(TechnicalFailureFlag = flag_colors)
    )
    Heatmap(
      matrix = x,
      col = cor_colors,
      name = "Correlation",
      left_annotation = left_annotation,
      top_annotation = top_annotation,
      ...
    )
  }
)
```

Let's try it out:

```{r}
plot(result)

# We can also pass options to `Heatmap()`:
plot(result, show_column_names = FALSE, show_row_names = FALSE)
```

