---
title: "Replace tern::df_explicit_na with a simpler hermes function"
author: "Haocheng Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Objective: replace `tern::df_explicit_na` with a simpler hermes function
 
For the example, we use the `colData` from the `summarized_experiment` object in the `hermes` package. We add one missing value to the categorical variable `AGE18`, add one missing value to the continuous variable `BAGE`, and one missing value to the logical variable `LowDepthFlag`.

```{r}
library(hermes)
dat <- colData(summarized_experiment)
dat[1, 4] <- NA
dat[1, 59] <- NA
dat[1, 86] <- NA
```

### Raw Code
We use `forcats::fct_explicit_na()` function to add a factor level `(Missing)`. This skips using the `df_explicit_na` functions in `tern`.

We use `lapply(, factor)` to convert character variables into factors without using loop, which is used in `tern::df_explicit_na`.

```{r}
hermes_explicit_na <- function(data, na_level = "<Missing>") {
    assert_that(is(data, "DataFrame"))
  
    var_is_logical <- sapply(data, is.logical)
    data[,var_is_logical] <- lapply(data[,var_is_logical], as.character)
      
    var_is_character <- sapply(data, is.character)
    data[,var_is_character] <- lapply(data[,var_is_character], factor)
    
    var_has_missing <- sapply(data, anyNA)
    var_to_add_NA_level <- var_is_character & var_has_missing
    data[,var_to_add_NA_level] <- lapply(data[,var_to_add_NA_level], 
                                         function(s) forcats::fct_explicit_na(s, 
                                                            na_level = na_level)       
                                         )
    data

}

```

### Test
A new level `(Missing)` is added to `AGE18`.
```{r}
dat_hermes_convert <- hermes_explicit_na(dat)

dat_hermes_convert$AGE18
```

Continuous variable `BAGE` does not have any change.
```{r}
dat_hermes_convert$BAGE
```

Logical variable is converted to factor. A new level `(Missing)` is added.
```{r}
dat_hermes_convert$LowDepthFlag
```

Original variable with all records missing is converted to a factor with one level `(Missing)`.
```{r}
dat_hermes_convert$RACE3
```
