---
title: "Design: PCA analysis and plot"
author: "Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This vignette explores the PCA analysis and plot.

## Idea

We want to replace [`draw_pca()`](https://github.roche.com/Rpackages/rnaseqTools/blob/dae3c425cf5bf0d45848323f722b558ce3bbf851/R/gSE_Metrics.R#L233).
- One function `calc_pca()` should do calculate the PCA
- the other function `draw_pca()` takes those results and create the plot (but see below...)
- The advantage of this modular design is that we can better test the PCA calculation and the user can also modify things / have their function before putting into the drawing function.

## Workflow

The workflow is expected as follows:
1. `HermesData` object created
2. QC flags added
3. Filtering
4. Normalization
5. PCA on the assay of interest (this can be original counts, CPM, log CPM, etc. which come
out of the normalization)
6. Plot

## Calculation Function

According to the workflow above, the PCA functions no longer need to calculate the CPM, log of it,
or any subsetting or filtering, as this has been done beforehand already.

However the user needs to be able to specify the assay to use.
We fix the scale and centering settings of the PCA analysis as they are advised. Since we scale,
we need to omit any genes that are constant across all samples.

```{r}
calc_pca <- function(object,
                     assay_name = "counts") {
  assert_that(
    is_hermes_data(object),
    is.string(assay_name)
  )
  
  # Obtain a matrix where each column is a gene, and keep only non-constant genes.
  x_samples <- assay(object, assay_name)
  x_genes <- t(x_samples)
  gene_is_constant <- apply(x_genes, MARGIN = 2L, FUN = isConstant)
  x_genes_remaining <- x_genes[, !gene_is_constant]
  
  stats::prcomp(
    x = x_genes_remaining,
    center = TRUE,
    scale = TRUE, 
    tol = sqrt(.Machine$double.eps)  # Omit essentially constant principal components.
  )
}

object <- HermesData(summarized_experiment)
pca <- calc_pca(object)
summary(pca)
```

## Plot Function

Now we can do the plot based on the PCA result.

### Evaluation of `ggfortify`

We could add this new package (which is already e.g. on BEE) to our docker containers.

```{r}
install.packages("ggfortify")
library(ggfortify)
autoplot(pca)
help(autoplot.prcomp)
```

So it seems this has quite a few options. See [here](https://rpubs.com/sinhrks/plot_pca) for more background. For example:

```{r}
# Different components.
autoplot(pca, x = 2, y = 4)

# Draw eigenvectors.
autoplot(pca, loadings = TRUE, loadings.label = TRUE)

# This fails, seems like a bug:
# autoplot(pca, loadings.data = pca$rotation[1:3, ], loadings = TRUE) 
```

Only issue here is that I have not found a good way so far to only plot selected eigenvectors here.

```{r}
# Sample labels.
autoplot(pca, label = TRUE, label.size = 2, label.hjust = 1, label.vjust = 1)

# Color by a factor in colData.
dta <- as.data.frame(colData(object))
autoplot(pca, data = dta, colour = "SEX")
```

So that works pretty nicely.

### Alternatives

Users could also go independently to [PCAtools](https://bioconductor.org/packages/release/bioc/html/PCAtools.html) starting from the 
normalized `HermesData` assays.

### Conclusion

Based on the easy use of `ggfortify`'s `autoplot`, we don't actually need to include dedicated
plot function ourselves in `hermes`. We can document in the examples how to use that, as well as in
the BAC.

