---
title: "Design: mapping arguments for HermesData constructor"
author: "Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Design: mapping arguments for HermesData constructor}
editor_options: 
  chunk_output_type: console
---

This vignette explores the introduction of mapping arguments to the `HermesData` constructor.

## Current state

The `HermesData` constructor currently just takes a `SummarizedExperiment` object and no
other arguments. First it checks for missing required columns in the `rowData` and `colData`
and fills in missing ones with `NA`s. Second it derives and checks the gene ID prefix, and then 
hands over to the internal default constructors which don't do anything besides validating the
object. The validation includes checks for the counts assay (e.g. it needs to be called `counts`),
the required columns and the names and gene ID prefix.

## Motivation

Often the user will have a `SummarizedExperiment` object with counts data, but there can be 
different names used:

- The assay with the raw counts might not be called `counts` but e.g. `count`.
- The required columns might have different names, e.g. `SYMBOL` instead of `symbol`, 
  `LowDepthFlag` instead of `low_depth_flag` etc.
  
Now the user could manually map the names to be aligned with what the constructor needs or to 
avoid duplication of columns internally (e.g. filling a `symbol` column with `NA`s just because
the `SYMBOL` column cannot be used). But that would be tedious. Therefore we want to make it easier
for the user to have this mapping been done as part of the constructor work.

## Arguments ideas

### One list argument

So the interface could look like this:

```{r, eval = FALSE}
HermesData(object, map_names = list(
  row_data = c(symbol = "HGNC"), 
  col_data = c(low_depth_flag = "LowDepthFlag"),
  assay = c(counts = "count")
))
```

This would be one `map_names` argument taking a list of three character vectors
`row_data`, `col_data` and `assay`, where the left-hand (name) side is the target
and the right-hand (value) side is the source.

If one of these vectors is `NULL` that is not specified in the list then nothing
will be renamed there.

We would not need to be strict about which targets are used, it would not need
to only be required names in `HermesData` objects but also renaming to other targets would 
be ok.

### Three separate arguments

We could also work with three separate arguments:

```{r, eval = FALSE}
HermesData(object, 
  row_data_map = c(symbol = "HGNC"), 
  col_data_map = c(low_depth_flag = "LowDepthFlag"),
  assay_map = c(counts = "count")
)
```

But that looks a bit too clunky.

### Implementation details

In the constructor, we would do the renaming as the first action, before checking
for required columns. Something like this:

```{r, message = FALSE}
library(hermes)
library(checkmate)

HermesData2 <- function(object, map_names = list()) {
  assert_that(
    is_class(object, "SummarizedExperiment"),
    not_empty(assays(object))
  )
  
  # Rename in separate function.
  object <- rename(object, map_names = map_names)
  
  # Other actions here.
  
  # Then return.
  .HermesData(object)
}
```

Actually, looking at it like this - we just want a specific `rename` method:

```{r}
h_map_pos <- function(names, map) {
  assert_character(map, min.chars = 1L, any.missing = FALSE, unique = TRUE, names = "unique")
  assert_subset(map, names)
  match(map, names)
}

setMethod(
  "rename", 
  signature = "SummarizedExperiment", 
  definition = function(x, mapping) {
    assert_list(mapping, types = "character")
    assert_subset(names(mapping), c("row_data", "col_data", "assay"))
    
    if (!is.null(row_map <- mapping$row_data)) {
      row_data <- rowData(x)
      col_names <- names(row_data)
      col_pos <- h_map_pos(names = col_names, map = row_map)
      names(rowData(x))[col_pos] <- names(row_map)
    }
    if (!is.null(col_map <- mapping$col_data)) {
      col_data <- colData(x)
      col_names <- names(col_data)
      col_pos <- h_map_pos(names = col_names, map = col_map)
      names(colData(x))[col_pos] <- names(col_map)
    }
    if (!is.null(assay_map <- mapping$assay)) {
      assay_names <- assayNames(x)
      assay_pos <- h_map_pos(names = assay_names, map = assay_map)
      assayNames(x)[assay_pos] <- names(assay_map)
    }
    x
  }
)
```

Let's try this out:

```{r}
object <- summarized_experiment
assayNames(object) <- "count"

object2 <- rename(object, mapping = list(
  row_data = c(symbol = "HGNC"), 
  col_data = c(low_depth_flag = "LowDepthFlag"),
  assay = c(counts = "count")
))

names(rowData(object2))
names(colData(object2))
assayNames(object2)
```

This is nice. The user can use this in the pipeline from the original input to the final
`HermesData` object in a very transparent manner. So we will not make this a constructor argument
but a separate rename method instead.
