---
title: "Box plot to accommodate by multiple genes"
author: "Namrata Bhatia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Objective: Implementation idea on adding the ability to specify multiple genes next to each other to the draw_boxplot()

## Original draw_boxplot()

Here we just have faceting and stratification by other `colData` variables.

```{r}
object <- HermesData(summarized_experiment)

assay_name = "counts"
x_var = "SEX"
y_var = genes(object)[19]
facet_var = "RACE"
color_var = "AGE18"

assay_matrix <- assay(object, assay_name)
col_data <- colData(object)
df <- data.frame(
  x = col_data[, x_var],
  y = assay_matrix[y_var, ]
) 
df$color <- col_data[[color_var]]
df$facet <- col_data[[facet_var]]


ggplot(df, aes(x = .data$x, y = .data$y)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar") +
  labs(x = x_var, y = y_var) +
  geom_point(aes(color = .data$color)) +
      labs(color = color_var) +
  facet_wrap(~facet)
```


## Update to add a way to specify multiple genes

### Option 1: for loop

We could do a for loop for `y_var` and keep `x_var` and `facet_var` to allow the user to be able to see distribution across different genes specified in the `y_var` by `x_var` and `facet_var`.

```{r}
y_vars = genes(object)[1:2]
facet_vars = genes(object)[1:2]

for (thisyvar in y_vars) {
  df <- data.frame(
    x = col_data[, x_var],
    y = assay_matrix[thisyvar, ]
  )
  df$color <- col_data[[color_var]]
  df$facet <- col_data[[facet_var]]
  
  print(ggplot(df, aes(x = .data$x, y = .data$y)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar") +
  labs(x = x_var, y = thisyvar) +
  geom_point(aes(color = .data$color)) +
      labs(color = color_var) +
  facet_wrap(~facet)
  )
}

# Test that this approach works when only one gene is specified
for (thisyvar in y_var) {
  df <- data.frame(
    x = col_data[, x_var],
    y = assay_matrix[thisyvar, ]
  )
  df$color <- col_data[[color_var]]
  df$facet <- col_data[[facet_var]]
  
  print(ggplot(df, aes(x = .data$x, y = .data$y)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar") +
  labs(x = x_var, y = thisyvar) +
  geom_point(aes(color = .data$color)) +
      labs(color = color_var) +
  facet_wrap(~facet)
  )
}
```

However, the problem here is that it will be difficult to compare within a facet the different gene distributions as it is too far away visually. Also for multiple genes there will be too many plots on the page.

### Option 2: Use `fill` aesthetic for the genes

The idea here is that we can add an additional aesthetic on top of the existing ones.
Here the `fill` aesthetic is a natural choice as it allows to plot the gene specific boxes next to each other.

In order to achieve this we need to use a long format representation of the different gene expression values.

```{r}
object <- HermesData(summarized_experiment)

assay_name <- "counts"
x_var <- "SEX"
y_var <- genes(object)[19]
y_vars <- genes(object)[1:2]
facet_var <- "RACE"
color_var <- "AGE18"

assay_matrix <- assay(object, assay_name)
col_data <- colData(object)
df2 <- data.frame(
  x = col_data[, x_var],
  color = col_data[, color_var],
  facet = col_data[, facet_var],
  y = as.numeric(t(assay_matrix[y_vars, , drop = FALSE])),
  fill = factor(rep(y_vars, each = ncol(assay_matrix)))
) 
```

Note that in the `data.frame()` call above, the three `colData` variables are repeated for each gene (as specified in `y_vars`). Therefore we need to transpose the subset of the assay matrix such that each gene has a column and the rows represent the samples. When we then coerce to a numeric vector the columns are bound together, we have all expression values for the first gene, then for the second gene, etc. This then matches the `colData` repetitions correctly. 

Given the long format it is almost as simple as before to plot the boxplots, as we just add the additional aesthetic.

```{r}
ggplot(df2, aes(x = .data$x, y = .data$y, fill = .data$fill)) +
  geom_boxplot() +
  stat_boxplot(geom = "errorbar") +
  geom_point(
    position = position_jitterdodge(jitter.width = 0),
    aes(color = .data$color, group = .data$fill)
  ) +
  labs(x = x_var, y = assay_name, color = color_var, fill = "Gene") +
  facet_wrap(~ facet)
```

We just need to make sure the points position uses the `position_jitterdodge()` function, see https://ggplot2.tidyverse.org/reference/position_jitterdodge.html. 
Depending on whether we want to jitter the points or not we can set `jitter_width` to 0 or leave at default (which uses jittering).

For the y-axis label we can show the assay name, since we show the gene ID(s) now in the legend of the plot.


