---
title: "Design: Adding quality control flags"
author: "Daniel Sabanes Bove"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
%\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---
  
This vignette prototypes the quality control flags addition functionality.

## Idea

After the user imports `HermesData` from a `SummarizedExperiment`, they still need to add 
certain quality control flag information to the object. 
- The settings are provided by a corresponding control function (which thresholds to use)
- Note that these columns must already be present in the object, but can be completely `NA` in the beginning.
- User interface is a function `add_quality_flags()`. No need for a generic here since this is quite
specific.
- Actual work is done by helper functions, one for each of the flags.

## Control function

This will import a function from the `tern` package for asserting proportion input.

```{r}
control_quality <- function(expression_min_cpm = 1,
                            expression_min_prop = 0.25,
                            technical_min_corr = 0.5,
                            depth_min = NULL) {
  assert_that(
    is.number(expression_min_cpm) && expression_min_cpm >= 0,
    expect_proportion(expression_min_prop),
    expect_proportion(technical_min_corr),
    is.null(depth_min) || is.number(depth_min)
  )  
  list(
    expression_min_cpm = expression_min_cpm,
    expression_min_prop = expression_min_prop,
    technical_min_corr = technical_min_corr,
    depth_min = depth_min
  )
}

cont <- control_quality()
```

## Helper functions

These do the actual work.

### `h_low_expression_flag()`

This calculates a low expression flag with one logical outcome per gene.

This function compares the CPM with a minimum expression CPM threshold. It counts for each gene how
many samples don't pass this threshold. If too many, then it flags this gene as "low expression" gene.

Note that we don't return any metadata, in contrast to the old function. This is because the control
information together with the returned logical flag variable provide all the information already (e.g.
number of genes total, excluded and included is easily obtained).

```{r}
h_low_expression_flag <- function(object, control) {
  assert_that(
    is_hermes_data(object),
    utils.nest::is_fully_named_list(control)
  )
  cpm <- edgeR::cpm(counts(object))
  threshold_n_samples <- ceiling(ncol(cpm) * control$expression_min_prop)
  n_samples_below_min_cpm <- rowSums(cpm <= control$expression_min_cpm)
  n_samples_below_min_cpm > threshold_n_samples
}

object <- hermes_data
res <- h_low_expression_flag(object, cont)
head(res)
```

### `h_tech_failure_flag()`

This calculates a technical failure flag with one logical outcome per sample.

This function first calculates the Pearson correlation matrix of the sample wise CPMs. So we have a matrix
that shows the correlation between samples. Then we compare the average correlation per sample with a threshold - if the average correlation is too low, the sample is flagged as "technical failure".

For now we don't return the correlation matrix. I think this is needed for PCA so we then get a dedicated function which we can reuse here.
  
```{r}
h_tech_failure_flag <- function(object, control) {
  assert_that(
    is_hermes_data(object),
    utils.nest::is_fully_named_list(control)
  )
  cpm <- edgeR::cpm(counts(object))
  sample_cor_matrix <- stats::cor(cpm, method = "pearson")
  average_cor_per_sample <- colMeans(sample_cor_matrix)
  average_cor_per_sample < control$technical_min_corr
}

h_tech_failure_flag(object, cont)
```

### `h_low_depth_flag()`

This calculates a low depth flag with one logical outcome per sample.

It computes the library size (total number of counts) per sample (removing any NAs).
If this number is too low, the sample is flagged as "low depth". By default,
this threshold is gives as the first quartile minus 1.5 times the inter quartile range. So anything
below the usual lower boxplot whisker would be too low.

```{r}
h_low_depth_flag <- function(object, control) {
  assert_that(
    is_hermes_data(object),
    utils.nest::is_fully_named_list(control)
  )
  lib_sizes <- colSums(counts(object))
  if (is.null(control$depth_min)) {
    lower_upper_quartiles <- quantile(lib_sizes, probs = c(0.25, 0.75))
    control$depth_min <- lower_upper_quartiles[1] - 1.5 * diff(lower_upper_quartiles)
  }
  lib_sizes < control$depth_min
}

h_low_depth_flag(object, cont)
```

## User function

Finally we get to the user interfacing function.

This sets the flags and saves the control options as metadata. In addition, it checks if the 
object already has such metadata. In that case, it only proceeds if the user wants to overwrite the
previous results.

```{r}
add_quality_flags <- function(object, 
                              control = control_quality(),
                              overwrite = FALSE) {
  assert_that(
    is_hermes_data(object),
    is.flag(overwrite)
  )
  already_added <- ("control_quality_flags" %in% names(metadata(object))) 
  if (already_added) {
    if (overwrite) {
      message("previously have added quality flags, but overwriting now")
    } else {
      stop("previously have added quality flags, please double check or ask for overwrite")
    }
  }
  
  rowData(object)$LowExpressionFlag <- h_low_expression_flag(object, control)
  colData(object)$TechnicalFailureFlag <- h_tech_failure_flag(object, control)
  colData(object)$LowDepthFlag <- h_low_depth_flag(object, control)
  
  metadata(object)$control_quality_flags <- control
  
  object
}

result <- add_quality_flags(object, cont)
result2 <- add_quality_flags(result, cont)
result2 <- add_quality_flags(result, cont, overwrite = TRUE)
```

